A Large-scale Study of Polymorphism in R

### 2019-03-31 paper/main.tex
ok

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left
}

@@

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}

@@

In this section, we will detail our methodology for collecting data.  
Basically, our aim is to observe arguments and return values of function calls, and from these generalize possible type signatures for the called functions. 
To this end, we build on an open source tool called \genthat whose purpose is to generate unit tests for R libraries~\cite{issta18}.  
\genthat achieves this by synthesizing unit tests from recorded function argument and return values, comparing the test against existing ones to avoid generating tests which do not increase code coverage.
To suit our purposes, we change the existing tool in two main ways: 
We record argument \emph{shapes} rather than values, and given the limited number of shapes this allowed us to avoid having to run \genthat's code coverage minimization phase. 
Both changes were beneficial for scalability, allowing us to trace far more calls than was reported in the \genthat work~\cite{issta18}.

\subsection{Example}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 1 + 2L                        # add double and integer
[1] 3
> x <- c(1,2,3,4)               # save double vector
> attr(x,"dim") <- c(2,2)       # assign to "dim" attribute
> x + c(1,2)                    # add double 3 to each matrix element matrix
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}\caption{Example script.}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
assignment to an attribute is written mixfix. The shapes we would expect to
record are as follows:

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
c: D[1], D[1], D[1], D[1] -> D[4]
`<-`: S, D[3] -> D[3]
c: D[1], D[1] -> D[2]
`attr<-1`:  S, C, D[2] -> 
c: D[1], D[1] -> D[2]
`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}

In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
\k{character}. For vectorized types we record their length and for all types
we record attributes with some of their values. Looking at the signatures
observed for addition, it is clear that the function is polymorphic as it
starts with the addition of two scalar numbers of different types, and then
adds a matrix to a vector returning a matrix.

@@

In R all arguments are passed as promises, and unused arguments will be unevaluated.
Our tracer does still collect information on these arguments, as whether or not they were used doesn't change the fact that they were in fact passed. 
In some cases, arguments aren't even intended to be evaluated.
For example, consider the {\tt magrittr} package, which implements the Tidyverse's pipe functions.
Refer to Figure~\ref{fig:magpipeex} for a translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}
Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)}, but that's erroneous code (you can't call {\tt mul\_by} without supplying two arguments).
In this situation, our tracer will still register the entire call, but flag the {\tt rhs} argument to indicate to later stages of the analysis that there was an issue evaluating that argument.

@@

@@

	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
	For example, \code{c(1, 2)[1.9] == 1}.

@@

	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.

@@

There are some clear limitations to the L0 type system.
As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
add_make_list <- function(x, y) {
  l <- as.list(x + y)
  attr(l, "example") <- "this is an attribute"
  l
}
add_make_list(2, c(3, 1)) # => list(5, 3)
\end{lstlisting}}
\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

@@

As for functions, roughly 22\% are polymorphic in type, or put differently 22\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 7742 & 9.66 \\ 
  scalar/double, scalar/integer & 7201 & 8.99 \\ 
  matrix, vector/double & 5384 & 6.72 \\ 
  NULL, scalar/character & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, scalar/double & 2082 & 2.60 \\ 
  NULL, vector/double & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, vector/character & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the FDTS.}
\end{table}

@@

In Table~\ref{tab:toppolyL1}, we see the prominence of {\tt double} and {\tt integer} polymorphism.
In many dynamic languages, primitive double and integer types can easily stand-in for each other as coercion between them is easy.
In R, the distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.
We address this by creating a new type, {\it real}, encompassing both.
The expected coercion is evident throughout R.
Refer to the snippet in Figure~\ref{fig:realex} for details.

As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
This is important as it shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.

Another clear pattern in Table~\ref{tab:toppolyFDTS} is that of NULL, X.
In many programming languages, NULLs inhabit every type, and this is not the case in our L1.
L1 reflects many of R's implementation details, and this distinction is no exception:
NULLs are represented as \AT{fill in} in the R runtime.

In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
Consider the snippet in Figure~\ref{fig:optnull} for an example.
\AT{TODO put that figure here}
The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.

Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
Just as we fold NULL into all types, we will do the same with NAs.

@@

>>>>

\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  framexleftmargin=-1mm,
  framexrightmargin=-1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
 % numbers=left
}

@@

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}

@@

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing the test against existing ones to avoid generating tests
which do not increase code coverage.  To suit our purposes, we change the
existing tool in two main ways: we record \emph{shapes} rather than values,
and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

@@

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L                     
[1] 3
> x <- c(1,2,3,4)            
> attr(x,"dim") <- c(2,2)    
> x + c(1,2)                 
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} & 
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+` &\tt D[1], I[1] -> D[1] \\
\tt c& D[1], D[1], D[1], D[1] -> D[4]\\
\tt `<-` &\tt S, D[3] -> D[3]\\
\tt c & \tt D[1], D[1] -> D[2]\\
\tt `attr<-1`&\tt  S, C, D[2] -> \\
\tt c &\tt D[1], D[1] -> D[2]\\
\tt `+` &\tt D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  The shapes we would expect to record are as follows:
In the above we abreviate types, \k{double}, \k{integer}, \k{symbol} and
\k{character}. For vectorized types we record their length and for all types
we record attributes with some of their values. Looking at the signatures
observed for addition, it is clear that the function is polymorphic as it
starts with the addition of two scalar numbers of different types, and then
adds a matrix to a vector returning a matrix.

@@

In R all arguments are passed as promises, and unused arguments will be
unevaluated.  Our tracer does still collect information on these arguments,
as whether or not they were used doesn't change the fact that they were in
fact passed.  In some cases, arguments aren't even intended to be evaluated.
For example, consider the {\tt magrittr} package, which implements the
Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}

Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
but that's erroneous code (you can't call {\tt mul\_by} without supplying
two arguments).  In this situation, our tracer will still register the
entire call, but flag the {\tt rhs} argument to indicate to later stages of
the analysis that there was an issue evaluating that argument.

@@

\AT{Maybe it's best to ignore attributes and classes for now?}

@@

	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
	For example, \code{c(1, 2)[1.9] == 1}.
	This distinction is mainly just relevant to the implementation, as integers and doubles are stored and dealt with differently in the runtime but programmers only see the distinction when printing.

@@

	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
	\AT{If we feel the need to go into more details on NULL or NA, I have the old paragraphs commented out in the source.}

@@

There are some clear limitations to the L0 type system that we should address before working through any of these other issues
As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
This is one of the many modifications we make, all of which are thoroughly explored in the following section.

%In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
%Consider the snippet in Figure~\ref{fig:optnull} for an example.
%\AT{TODO put that figure here}
%The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
%Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
%Just as we fold NULL into all types, we will do the same with NAs.

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
add_make_list <- function(x, y) {
  l <- as.list(x + y)
  attr(l, "example") <- "this is an attribute"
  l
}
add_make_list(2, c(3, 1)) # => list(5, 3)
\end{lstlisting}}

\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

@@

As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 7742 & 9.66 \\ 
  scalar/double, scalar/integer & 7201 & 8.99 \\ 
  matrix, vector/double & 5384 & 6.72 \\ 
  NULL, scalar/character & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, scalar/double & 2082 & 2.60 \\ 
  NULL, vector/double & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, vector/character & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the L1.}
\end{table}

@@

In Table~\ref{tab:toppolyL1}, we again see the prominence of {\tt double} and {\tt integer} polymorphism.
As discussed previously, we will address this by creating a new {\it real} type encompassing both.
Refer back to the snippet in Figure~\ref{fig:realex} for details.

As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.

Another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
As we discussed, we will roll NULL (and raw NAs) into other types.

@@

%
%
%
%
\subsection{Metaprogramming}

\AT{Will probably change.}

Tidyverse owes some of its success to the new syntax defined by some of the included packages.
One such example is the pipe operator \%$>$\% that we saw in the last section, and indeed this function was highly exercised by many other packages \AT{get counts}.
We also grappled with the unfortunate truth that arguments to \%$>$\% can't always be evaluated, which causes some annoyances in the analysis.

We would like to know how often packages contain functions designed for metaprogramming.
As it happens, when arguments fail to evaluate during our analysis even as functions execute successfully, we are actually dealing with instances of metaprogramming.
Some functions, such as those in the magrittr package, will take one (or more) of their arguments and use them as an AST, performing surgery on the expression to create some new expression that they evaluate.
Sometimes, these arguments intended for metaprogramming {\it will} successfully evaluate in our tracer, but this is merely chance.
For example, consider the code snippet in Figure~\ref{fig:metaexworks}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by_w_defaults <- function(x=1, y=1) x * y

4 %>% mul_by_w_defaults(2) == `%>%`(lhs = 4, rhs = mul_by(3)) 
\end{lstlisting}}\caption{Dates in R.}\label{fig:metaexworks}\end{figure}

In this example, our analysis will successfully evaluate the {\tt rhs} argument
